---
title: Biostatistics 140.623 
subtitle: <h3 style="font-style:normal">Third Term, 2017-2018 <br> Problem Set 2 (with R)</h3>
author: <h4 style="font-style:normal">Martin Skarzynski</h4>
date: <h4 style="font-style:normal">02/15/2018</h4>
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Biostatistics 140.623

Problem Set 2

Biostatistics 140.623
Third Term, 2017-2018
Problem Set 2 (with R)
Survival in Primary Biliary Cirrhosis
Learning Objectives:
Students who successfully complete this section will be able to:
- To evaluate whether the drug DPCA prolongs life in patients.
- To identify baseline characteristics of patients which predict longer survival.
- Analyze the survival time data (without grouping) by the Kaplan-Meier estimate of the
survival function, the log- rank statistic, and Cox proportional hazards model.
- Check the estimated model for its consistency with the observed data; in particular, check
the proportional hazards assumption using the complementary log-log plot of the
estimated survival function.
- Summarize the findings for public health readers and document and archive the steps of
the statistical analysis by creating a script file in R.

Data Set:
Between January 1974 and May 1984, a double-blinded randomized trial on patients with primary biliary cirrhosis (PBC) of the liver was conducted at the Mayo clinic. A total of 312 patients were randomized to either receive the drug D-penicillin (DPCA) or a placebo. Patients were followed until they died from PBC or until censoring, either because of administrative censoring (withdrawn alive at end of study), death not attributable to PBC, liver transplantation, or loss to follow-up. At baseline, a large number of clinical, biochemical, serological and histologic measurements were recorded on each patient. This data set is a subset of the original data, and includes information on each patient’s time to death or censoring, treatment, age, gender, serum bilirubin, and histologic disease stage (1-4).
The variables included in this dataset include:
case: unique patient ID number
sex: 0 = male, 1 = female (coded as "Female" and "Male" in the csv file rather than 0/1)
drug: 0 = placebo, 1 = DPCA
bil : serum bilirubin in mg/dl
survyr: time (in years) to death or censoring
death: indicator = 1 if patient died, 0 if censored
ageyr: age in years [continuous variable]
histo: histologic disease stage (1 – 4) [categorical variable]
agecat: age categories, coded as "< 45 yrs", "45 – 55 yrs", and ">= 55 yrs"
Also included in the data set for your possible use are the following indicator (dummy) variables:

Age Indicators (indicator versions of agecat):
agegr_2: 1 if patient is 45-55 years old, 0 otherwise
agegr_3: 1 if patient is >= 55 years old, 0 otherwise
Histologic Stage Indicators:
hstage2: 1 if patient is in Stage 2, 0 otherwise
hstage3: 1 if patient is in Stage 3, 0 otherwise
hstage4: 1 if patient is in Stage 4, 0 otherwise
The data are stored in the csv data set pbctrial.csv, which may be
downloaded from the course website.
Methods:
Use the data set described above and the appropriate statistical analyses to address the specific
learning objectives listed on the first page.
Hints: The hints shown below are based on a dataset with the name pbcData, read in with the
following code. In the following list of commands, if you want to look at differences by other
variables than drug, you should change the variable name! Create a new .R file to type/run your
commands so that you will have a record of your analysis.
```{r}

library(readr)
pbcData = read_csv("pbctrial.csv")
```


a. Explore the data using descriptive statistics:
table()
prop.table()
summary()
etc

```{r}

dim(pbcData)
str(pbcData)
summary(pbcData)
```

b. Define a survival object, defining the time variable (survyr) and the event (death == 1).
To do this, you must first install and load the "survival" package:
```{r}

# install.packages("survival")
library(survival)
```

## only run this the first time

```{r}
pbcData$SurvObj = with(pbcData, Surv(survyr, death == 1))

```

c. Explore differences in time to death by different baseline variables using graphs and
complementary log-log plots.
```{r}

# estimate survival curves for entire sample
km.overall = survfit(SurvObj ~ 1, data = pbcData,
type="kaplan-meier", conf.type="log-log")
km.overall
summary(km.overall)
# estimate survival curves for drug group
km.drug = survfit(SurvObj ~ drug, data = pbcData,
type="kaplan-meier", conf.type="log-log")
km.drug
summary(km.drug)
```

```{r}

# plot km curves
plot(km.overall)
plot(km.drug)
# log rank test for equality of survivor functions
survdiff(SurvObj ~ drug, data=pbcData)
# complimentary log-log plot
plot(km.drug, fun="cloglog", ylab="log(-log(Survival Probability)",
xlab="Analysis time (shown on log scale)")
```

d. Fit several Cox proportional hazards regression models to the ungrouped survival data:
```{r}

model1 = coxph(SurvObj ~ drug, data = pbcData)
summary(model1)

model2 = coxph(SurvObj ~ sex + bil + as.factor(histo), data = pbcData)
summary(model2)
```

e. Save your R script file that documents and archives the steps of your statistical analysis.
This file will make your analysis "reproducible."
f. Summarize your findings in a brief report (less than two pages with at most one table and
one figure) as if for a biomedical/public health journal.
A suggested format is:
- Introduction – a few sentences about the research question(s)
- Data description – simple tabulations describing patient characteristics
- Results from multiple models that address question(s) (e.g., bivariate and multivariable)
- Graphical display that presents evidence in the data relevant to your scientific question.

## Introduction 
The research question that I will try to answer in this report is whether D-penicillin (DPCA), the drug tested in the PBC trial, provided any benefit for the patient population as a whole (n=312) and for sub-groups based on sex, age and disease stage. I hypothesize that the drug effect will not be different between the 3 age categories, but will depend on disease stage. In other wrods, I expect that there will be differences in time to death between the 4 disease stages, specifically that more advanced disease will be more difficult to treat, which will result in a shorter time to event. I will also assess whether bilirubin is a prognostic marker and whether drug beenfit will differ among men versus women.

## Data description 
The best way to describe the data in my opinion is using the `skim` function from the `skimr` R package. This function produces a table of descriptive statistics but also small histograms showing the distribution of each variable. 
```{r tab}
# install.packages("skimr")
library(skimr)
skimr::skim(pbcData)
```
There are a total of 312 patients and the median survival time was around 5 years. As for patient characteristics, the representation across age categories and disease stages appears to spread relatively evenly. The `age` and `survyr` variable appear to be normally distributed with a slight leftward skew. Interestingly, bilirubin is skewed highly to the left indicating that there are outliers with high bilirubin values.

## Results 
First, I decided to put all variables of interest into one model rather creating multiple models that address each of the above questions, because the instructions say to have at most one figure and one table. If any of the results are statistically significant, I can explore the question further with a more specific model in the future. First some basic exploratory data analysis will let me know if I am on the right track with the variables I have chosen. If there is no difference between the median survival times of the groups I am interested in, it will be unlikely that I will see anything significant in my model.
```{r mod3}
model3 = coxph(SurvObj ~ sex + bil + as.factor(histo) + as.factor(agecat), data = pbcData)
summary(model3)
```

```{r eda1}
model3 = coxph(SurvObj ~ sex + bil + as.factor(histo) + as.factor(agecat), data = pbcData)
summary(model3)
library(dplyr)
# install.packages("broom")
library(broom)
pbcData %>%
    group_by(sex, drug) %>%
    summarise(med_surv = median(survyr))
pbcData %>%
    group_by(agecat, drug) %>%
    summarise(med_surv = median(survyr))
pbcData %>%
    group_by(histo, drug) %>%
    summarise(med_surv = median(survyr))
```
From this initial analysis it looks like patients in the highest age category that were given placebo fare the worst. These results indicate that elderly patients my stand to benefit the most from taking the drug. Shockingly, men taking the drug appear to have a shorter survival time than with the drug, and do not survive as long as a women in general. Similarly, the drug appeared to have a negative effect on survival in patients with earliest stage of disease (histo = 1). Now I will take a similar approach to the data but using a cox model.
```{r eda2}
glance(model3)
aug <- augment(model3, pbcData)
risk_aug <- augment(model3, pbcData, type.predict = "risk")
exp_aug <- augment(model3, pbcData, type.predict = "expected")

par(mfrow=c(3,3))
## install.packages("ggplot2")
library(ggplot2)
ggplot(aug, aes(ageyr, .fitted, color = sex, shape = as.factor(histo))) + geom_point()
ggplot(risk_aug, aes(ageyr, .fitted, color = sex, shape = as.factor(histo))) + geom_point()
ggplot(exp_aug, aes(ageyr, .fitted, color = sex, shape = as.factor(histo))) + geom_point()

ggplot(aug, aes(bil, .fitted, color = sex, shape = as.factor(histo))) + geom_point()
ggplot(risk_aug, aes(bil, .fitted, color = sex, shape = as.factor(histo))) + geom_point()
ggplot(exp_aug, aes(bil, .fitted, color = sex, shape = as.factor(histo))) + geom_point()
```

## Graphical display 
I plotted cox model fitted values against ageyr and bil variables marking sex and disease stage (`histo`) with color and different symbols, respectively. 

## Conclusions

It is clear that all of the variables I picked are important in the final model although not all levels of the categorical variables were statistically significant. This work is only the beginning and more precise answers to the research questions discussed in the introduction will require further inspection with models more precisely adapted to each research question.
